---
title: "Spatial Modeling of Collisions in NYC"
author: "Peter Silverstein"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
# Libraries
library(tidyverse)
library(sf)
library(tidycensus)
library(keyring)
library(DT)
library(tmap)
library(spdep)
```

# Introduction

In a prior post (LINK HERE), I began this project: a bit of a personal workshop in learning how to use Stan for spatial modeling. In that post, I did some exploratory data analysis to set up the overarching idea of the project, which is to (a) follow the example in LINK PAPER and use an ICAR model to model spatial dependencies in crashes across NYC and (b) potentially add a treatment/difference in difference component to this model to understand whether the congestion pricing policy has an effect on the number of vehicle crashes. In the interest of breaking this project into digestible chunks, I have decided that this, part 2 of 3, will focus only on the spatial modeling part. To that end, I'll be basically following this GUIDE from Mitzi Morris.

```{r class.source=NULL, message=FALSE, warning=FALSE}
# Loading NYC Census Tracts: https://data.cityofnewyork.us/City-Government/2020-Census-Tracts/63ge-mke6/about_data 
nyc_tracts <- read.csv("data/2020_Census_Tracts_20250606.csv", stringsAsFactors = FALSE) %>%
  rename("geometry" = "the_geom")
nyc_tracts <- st_as_sf(nyc_tracts, wkt = "geometry", crs = 4326)
nyc_tracts <- nyc_tracts %>%
  mutate(
    area_sqm = Shape_Area / 27878400
  ) %>%
  select(
    geometry,
    BoroCT2020,
    BoroCode,
    BoroName,
    area_sqm,
    GEOID
  )

# Loading collisions data: https://data.cityofnewyork.us/Public-Safety/Motor-Vehicle-Collisions-Crashes/h9gi-nx95/about_data 
filter <- c(0, NA)

crashes <- read.csv("data/Motor_Vehicle_Collisions_-_Crashes_20250605.csv") %>%
  select(CRASH.DATE,
         LATITUDE,
         LONGITUDE,
         NUMBER.OF.PERSONS.INJURED,
         NUMBER.OF.PERSONS.KILLED,
         NUMBER.OF.PEDESTRIANS.INJURED,
         NUMBER.OF.PEDESTRIANS.KILLED) %>%
  filter(! LATITUDE %in% filter,
         ! LONGITUDE %in% filter) %>%
  rename(
    "date" = "CRASH.DATE",
    "lat" = "LATITUDE",
    "long" = "LONGITUDE",
    "persons_inj" = "NUMBER.OF.PERSONS.INJURED",
    "persons_death" = "NUMBER.OF.PERSONS.KILLED",
    "ped_inj" = "NUMBER.OF.PEDESTRIANS.INJURED",
    "ped_death" = "NUMBER.OF.PEDESTRIANS.KILLED"
  ) %>%
  mutate(
    date = mdy(date)
    ) %>%
  st_as_sf(coords = c("long","lat"), crs = 4326)

# Grabbing some census variables via Tidycensus, using Keyring for API Key privacy
tidycensus_api_key <- key_get(service = "tidycensus_API", username = "my_tidycensus")
census_api_key(tidycensus_api_key)

census_vars <- get_acs(state = "NY",
                       county = c("Bronx", "Kings", "New York", "Queens", "Richmond"),
                       geography = "tract",
                       variables = c(medincome = "B19013_001",
                                     population = "B01003_001",
                                     median_age = "B01002_001",
                                     transport_baseline = "B08301_001",
                                     transport_pubtransit = "B08301_010"),
                       geometry = FALSE,
                       keep_geo_vars = FALSE,
                       year = 2022,
                       output = "wide"
                       ) %>%
  mutate(
    GEOID = as.numeric(GEOID),
    median_income = medincomeE,
    population = populationE,
    median_age = median_ageE,
    prop_pubtransit = transport_pubtransitE / transport_baselineE
  ) %>%
  select(
    GEOID,
    median_income,
    population,
    median_age,
    prop_pubtransit
  )

# Associating CP Zone, Pre/Post, Treatment, Borough, and Census Tract w/ Observations
crashes <- crashes %>%
  st_join(nyc_tracts, join = st_within) %>%
  filter(! is.na(area_sqm))

# Aggregating/summarizing data to census tract level, no time component
crashes_tract <- crashes %>%
  group_by(BoroCT2020) %>%
  summarize(tot_crashes = n(),
            area = mean(area_sqm)) %>%
  ungroup() %>%
  st_drop_geometry()

# Getting Fragmentation Index from: https://github.com/mitzimorris/geomed_2024/blob/main/data/nyc_study.geojson

frag_data = st_read(file.path("data", "nyc_study.geojson"), quiet = TRUE) %>%
  st_drop_geometry() %>%
  select(BoroCT2010, frag_index, traffic) %>%
  mutate(
    BoroCT2010 = as.numeric(BoroCT2010)
  )

# Joining everything together, selecting only variables that I want
crashes_tracts_geo <- nyc_tracts %>%
  right_join(crashes_tract, 
             by = "BoroCT2020") %>%
  left_join(census_vars,
            by = "GEOID") %>%
  left_join(frag_data,
            by = c("BoroCT2020" = "BoroCT2010")) %>%
  mutate(
    pop_per_sqm = population / area_sqm
  ) %>%
  select(! c(area, population))

# Interactive Data Table for Display
crashes_dt <- crashes_tracts_geo %>%
  st_drop_geometry() %>%
  mutate(
    area_sqm = round(area_sqm, 3),
    prop_pubtransit = round(prop_pubtransit, 3),
    pop_per_sqm = round(pop_per_sqm, 3),
  )

datatable(crashes_dt,
          extensions = 'Buttons',
          filter = "top",
          rownames = FALSE,
          options = list(
            autoWidth = TRUE,
            scrollX = TRUE
          ),
  class = 'compact',
  escape = FALSE
) %>%
  formatStyle(
    columns = names(crashes_dt),
    `white-space` = "nowrap",
    `height` = "1.5em",
    `line-height` = "1.5em"
  )
```

# More Plots

## Histogram of Collisions per Square Mile
```{r class.source=NULL}
ggplot(data = crashes_tracts_geo, aes(x = tot_crashes)) + 
  geom_histogram(aes(y = after_stat(density)), 
                 bins = 500, 
                 color = "lightblue",
                 fill = "lightblue",
                 alpha = 0.75) + 
  geom_density(color = "#FFC600", linewidth = 1) + 
  theme_bw() + 
  labs(title = "Distribution Total Crashes per Census Tract (2024-25)",
       x = "Number of Crashes",
       y = "Density")
```

## Choropleth Maps

### Map of Collisions
```{r}
crash_map <- tm_shape(crashes_tracts_geo) + 
  tm_polygons(fill = "tot_crashes",
              fill.scale = tm_scale_intervals(values = "brewer.yl_or_rd",
                                              style = "jenks"),
              fill.legend = tm_legend(title = "Total Crashes, 2024-25"),
              lwd = 0.3) +
  tm_layout(legend.outside = TRUE, 
            legend.outside.position = "right")

crash_map
```

### Predictor Maps

#### Median Income // Median Age
```{r}
income_map <- tm_shape(crashes_tracts_geo) + 
  tm_polygons(fill = "median_income",
              fill.scale = tm_scale_intervals(values = "brewer.greens",
                                              style = "jenks"),
              fill.legend = tm_legend(title = "Median Income"),
              lwd = 0.3) +
  tm_layout(legend.outside = TRUE, 
            legend.outside.position = "right")

medianage_map <- tm_shape(crashes_tracts_geo) + 
  tm_polygons(fill = "median_age",
              fill.scale = tm_scale_intervals(values = "brewer.blues",
                                              style = "jenks"),
              fill.legend = tm_legend(title = "Median Age"),
              lwd = 0.3) +
  tm_layout(legend.outside = TRUE, 
            legend.outside.position = "right")

tmap_arrange(income_map,
             medianage_map,
             nrow = 1, ncol = 2)
```

#### Proportion Using Public Transit to Commute // Population per Square Mile
```{r}
prop_pubtransit_map <- tm_shape(crashes_tracts_geo) + 
  tm_polygons(fill = "prop_pubtransit",
              fill.scale = tm_scale_intervals(values = "brewer.purples",
                                              style = "jenks",
                                              value.na = "grey"),
              fill.legend = tm_legend(title = "Proportion that Commute Using Public Transit"),
              lwd = 0.3) +
  tm_layout(legend.outside = TRUE, 
            legend.outside.position = "right")

pop_per_sqm_map <- tm_shape(crashes_tracts_geo) + 
  tm_polygons(fill = "pop_per_sqm",
              fill.scale = tm_scale_intervals(values = "brewer.yl_gn",
                                              style = "jenks"),
              fill.legend = tm_legend(title = "Population per Square Mile"),
              lwd = 0.3) +
  tm_layout(legend.outside = TRUE, 
            legend.outside.position = "right")

tmap_arrange(prop_pubtransit_map,
             pop_per_sqm_map,
             nrow = 1, ncol = 2)
```

#### Fragmentation Index // Traffic
```{r}
frag_map <- tm_shape(crashes_tracts_geo) + 
  tm_polygons(fill = "frag_index",
              fill.scale = tm_scale_intervals(values = "brewer.pu_or",
                                              style = "quantile",
                                              midpoint = 0,
                                              value.na = "grey"),
              fill.legend = tm_legend(title = "Fragmentation Index"),
              lwd = 0.3) +
  tm_layout(legend.outside = TRUE, 
            legend.outside.position = "right")

traffic_map <- tm_shape(crashes_tracts_geo) + 
  tm_polygons(fill = "traffic",
              fill.scale = tm_scale_intervals(values = "brewer.oranges",
                                              style = "quantile",
                                              midpoint = 0,
                                              value.na = "grey"),
              fill.legend = tm_legend(title = "Traffic"),
              lwd = 0.3) +
  tm_layout(legend.outside = TRUE, 
            legend.outside.position = "right")

tmap_arrange(frag_map,
             traffic_map,
             nrow = 1, ncol = 2)
```

# Moving from a Map to Spatial Dependencies

## Map/Graph of Neighbor Links
```{r warning=FALSE}
nyc_nbs = poly2nb(crashes_tracts_geo)
nyc_coords = st_coordinates(st_centroid(crashes_tracts_geo['geometry']))

# Create a spatial points sf object for centroids
centroids_sf <- st_as_sf(data.frame(x = nyc_coords[,1], y = nyc_coords[,2]), 
                         coords = c("x", "y"), 
                         crs = st_crs(crashes_tracts_geo))

# Plot with tmap
spdep_map <- tm_shape(crashes_tracts_geo) +
  tm_polygons(col = "white") + 
  tm_lines(col = "white", lwd = 2) + 

  # Add points for centroids
  tm_shape(centroids_sf) +
  tm_dots(size = 0.25, col = "black") +

  # Add lines for neighbors
  tm_shape(st_as_sf(nb2lines(nyc_nbs, coords = nyc_coords))) +
  tm_lines(col = "deepskyblue3")

spdep_map
```

# NEXT STEPS IS UNDERSTAND WHAT MORRIS IS DOING LOL





