---
title: "Causal Analysis of the Effect of Congestion Pricing on Traffic Accidents in NYC"
author: "Peter Silverstein"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
# Libraries
library(tidyverse)
library(sf)
library(tidycensus)
library(keyring)
library(DT)
library(tmap)
library(spdep)
library(cmdstanr)
library(posterior)
library(bayesplot)
library(loo)
library(parallel)
library(patchwork)
cores = floor(parallel::detectCores() / 2)
source("helper-functions/utils_dataviz.R")
source("helper-functions/utils_bym2.R")

# For use in opening file and saving time
# time_fit_1 <- readRDS("model-files/time_fit_1.rds")
# time_fit_2 <- readRDS("model-files/time_fit_2.rds")
# full_fit_1 <- readRDS("model-files/full_fit_1.rds")
```

# Introduction

This post is Part III of my project using Stan to model crash counts in New York City census tracts using the <a href="https://data.cityofnewyork.us/Public-Safety/Motor-Vehicle-Collisions-Crashes/h9gi-nx95/about_data"  target="_blank">Motor Vehicle Collisions - Crashes</a> dataset from NYC Open Data. The overarching goals of this project are to (a) improve my Stan modeling skill in a Bayesian workflow setting and (b) perform some causal inference to see if the <a href="https://congestionreliefzone.mta.info"  target="_blank">Congestion Relief Zone (CRZ)</a>, introduced in early 2025, impacted the number of collisions that occur in a census tract. In <a href="nyc-collisions-EDA.html"  target="_blank">Part I</a>, I performed exploratory data analysis through tables, visualizations, and tests for statistical clustering. In <a href="nyc-collisions-modeling.html"  target="_blank">Part II</a>, I modeled spatial correlation between crash counts in neighboring census tracts using a BYM2 model. Here, in Part III, I will extend the BYM2 model to include a time component, which means that my models will adjust for (a) unmeasured spatial variation, (b) unmeasured temporal variation, and (c) pre-treatment demographic and built-environment variables. This provides a good backbone for my eventual inclusion of a treatment indicator (whether the measurement was within the CRZ after congestion pricing was implemented), and an estimation of the treatment effect.

It is my intention to write a final post which presents the findings of this project in a concise manner. Knowing that the concise document is forthcoming, I can disregard concerns about length in this post and simply try to do good, transparent statistical modeling. In that vein, you can find the full GitHub repository for this project here: <a href="https://github.com/Peter-Silverstein/nyc-collisions"  target="_blank">https://github.com/Peter-Silverstein/nyc-collisions</a>.

A final note before I get into it: this is a personal passion project, my time is somewhat limited, and my computer has its limits. I think, with sophisticated Stan modeling, there's always another thing you can do. I want to get this done and posted, so my plan is to do quite a few of those things, find a good place to call the model decent, and interpret the results with that caveat.

# Data Preparation

The data preparation for this part is largely the same as the previous part, with two big exceptions. First, data are aggregated to the Tract/Month/Year level, to provide the hierarchical structure that my model needs. Second, I have included an indicator variable for whether the observed tract is within the congestion pricing zone and, for those that are, an indicator for whether the observed point is 01/2025 or later. I consider a tract observation within the zone from 01/2025 or later to be "treated."

```{r class.source=NULL, message=FALSE, warning=FALSE}
# Loading NYC Census Tracts: https://data.cityofnewyork.us/City-Government/2020-Census-Tracts/63ge-mke6/about_data 
nyc_tracts <- read.csv("data/2020_Census_Tracts_20250606.csv", stringsAsFactors = FALSE) %>%
  rename("geometry" = "the_geom")
nyc_tracts <- st_as_sf(nyc_tracts, wkt = "geometry", crs = 4326)
nyc_tracts <- nyc_tracts %>%
  mutate(
    area_sqm = Shape_Area / 27878400
  ) %>%
  select(
    geometry,
    BoroCT2020,
    BoroCode,
    BoroName,
    area_sqm,
    GEOID
  )

# Loading collisions data: https://data.cityofnewyork.us/Public-Safety/Motor-Vehicle-Collisions-Crashes/h9gi-nx95/about_data 
filter <- c(0, NA)

crashes <- read.csv("data/Motor_Vehicle_Collisions_-_Crashes_20250708.csv") %>%
  select(CRASH.DATE,
         LATITUDE,
         LONGITUDE,
         NUMBER.OF.PERSONS.INJURED,
         NUMBER.OF.PERSONS.KILLED,
         NUMBER.OF.PEDESTRIANS.INJURED,
         NUMBER.OF.PEDESTRIANS.KILLED) %>%
  filter(! LATITUDE %in% filter,
         ! LONGITUDE %in% filter) %>%
  rename(
    "date" = "CRASH.DATE",
    "lat" = "LATITUDE",
    "long" = "LONGITUDE",
    "persons_inj" = "NUMBER.OF.PERSONS.INJURED",
    "persons_death" = "NUMBER.OF.PERSONS.KILLED",
    "ped_inj" = "NUMBER.OF.PEDESTRIANS.INJURED",
    "ped_death" = "NUMBER.OF.PEDESTRIANS.KILLED"
  ) %>%
  mutate(
    date = mdy(date)
    ) %>%
  st_as_sf(coords = c("long","lat"), crs = 4326)

# Grabbing some census variables via Tidycensus, using Keyring for API Key privacy
tidycensus_api_key <- key_get(service = "tidycensus_API", username = "my_tidycensus")
census_api_key(tidycensus_api_key)

census_vars <- get_acs(state = "NY",
                       county = c("Bronx", "Kings", "New York", "Queens", "Richmond"),
                       geography = "tract",
                       variables = c(medincome = "B19013_001",
                                     population = "B01003_001",
                                     median_age = "B01002_001",
                                     transport_baseline = "B08301_001",
                                     transport_pubtransit = "B08301_010"),
                       geometry = FALSE,
                       keep_geo_vars = FALSE,
                       year = 2022,
                       output = "wide"
                       ) %>%
  mutate(
    GEOID = as.numeric(GEOID),
    median_income = medincomeE,
    population = populationE,
    median_age = median_ageE,
    prop_pubtransit = transport_pubtransitE / transport_baselineE
  ) %>%
  select(
    GEOID,
    median_income,
    population,
    median_age,
    prop_pubtransit
  )

# Associating Borough, and Census Tract w/ Observations
crashes <- crashes %>%
  st_join(nyc_tracts, join = st_within) %>%
  filter(! is.na(area_sqm))

# Loading Central Business District Shape: https://data.ny.gov/Transportation/MTA-Central-Business-District-Geofence-Beginning-J/srxy-5nxn/about_data
cbd_geofence <- read.csv("data/MTA_Central_Business_District_Geofence__Beginning_June_2024_20250605.csv", stringsAsFactors = FALSE)
cbd_geofence <- st_as_sfc(cbd_geofence$polygon, crs = 4326)

# Aggregating/summarizing data to census tract, month, year levels
# Note the included components to fill month/year/tract combination with no crashes with 0
crashes_tract <- crashes %>%
  mutate(
    monthyear = format(date, "%m-%y")
  ) %>%
  group_by(BoroCT2020, monthyear) %>%
  summarize(tot_crashes = n(),
            area = mean(area_sqm),
            .groups = "drop") %>%
  complete(BoroCT2020, monthyear, fill = list(tot_crashes = 0)) %>% 
  group_by(BoroCT2020) %>%
  fill(area, .direction = "downup") %>%
  ungroup() %>%
  st_drop_geometry() %>%
  select(! geometry) %>%
  mutate(
    month = as.numeric(substr(monthyear, 1, 2)),
    year = as.numeric(substr(monthyear, 4, 5))
  ) %>%
  filter(! (year == 25 & month > 6))

# Getting Fragmentation Index from: https://github.com/mitzimorris/geomed_2024/blob/main/data/nyc_study.geojson

frag_data = st_read(file.path("data", "nyc_study.geojson"), quiet = TRUE) %>%
  st_drop_geometry() %>%
  select(BoroCT2010, frag_index, traffic) %>%
  mutate(
    BoroCT2010 = as.numeric(BoroCT2010)
  )

# Joining everything together, selecting only variables that I want
crashes_tracts_geo <- nyc_tracts %>%
  right_join(crashes_tract, 
             by = "BoroCT2020") %>%
  left_join(census_vars,
            by = "GEOID") %>%
  left_join(frag_data,
            by = c("BoroCT2020" = "BoroCT2010")) %>%
  select(! c(area)) %>% 
  filter(! if_any(everything(), is.na)) %>%
  mutate(
    cp_zone = as.integer(lengths(st_intersects(geometry, cbd_geofence)) > 0),
    after_cp = ifelse(year == 25, 1, 0),
    treatment = ifelse(cp_zone == 1 & after_cp == 1, 1, 0)
  ) %>%
  mutate(
    tract_id = as.integer(factor(BoroCT2020)),
    time_id = as.integer(factor(monthyear))
  )

# Interactive Data Table for Display
crashes_dt <- crashes_tracts_geo %>%
  st_drop_geometry() %>%
  mutate(
    area_sqm = round(area_sqm, 3),
    prop_pubtransit = round(prop_pubtransit, 3)
  )

datatable(crashes_dt,
          extensions = 'Buttons',
          filter = "top",
          rownames = FALSE,
          options = list(
            autoWidth = TRUE,
            scrollX = TRUE
          ),
  class = 'compact',
  escape = FALSE
) %>%
  formatStyle(
    columns = names(crashes_dt),
    `white-space` = "nowrap",
    `height` = "1.5em",
    `line-height` = "1.5em"
  )
```

## Visualizing the Distribution of Counts

One thing that will become important later on is the shape and characteristics of the distribution of crash counts, so I visualize it for posterity here.

```{r class.source=NULL, message=FALSE, warning=FALSE}
ggplot(data = crashes_tracts_geo, aes(x = tot_crashes)) + 
  geom_histogram(aes(y = after_stat(density)), 
                 bins = 100, 
                 color = "lightblue",
                 fill = "lightblue",
                 alpha = 0.75) + 
  theme_bw() + 
  labs(title = "Distribution of Total Crashes per Census Tract (2024-25)",
       x = "Number of Crashes",
       y = "Density")
```

# Time Series Modeling

Given that I already built a working BYM2 spatial dependency model in the previous part, I'll start this time with a simple hierarchical model providing heterogeneous effects for each time period within my data range, January 2024 through May 2025. I doubt that the number of crashes in one month predicts the number of crashes in the next month (i.e., is the city really implementing policy on the fly to address minor variation in car crash counts?), but I do think that it's likely that crash counts vary by month (and the EDA in Part I supported this). To that end, the model below models spatial variation as a hierarchy within each time frame. The use of the simplex for rho is unnecessary at this stage, but I will be using it when I combine the spatial model with this one, so might as well get it implemented early.

## Initial Time Series Poisson Model

```{r class.source=NULL, message=FALSE, warning=FALSE}
# The Stan Model
time_model_1_file = file.path('stan', 'poisson_time_1.stan')
cat(readLines(time_model_1_file), sep="\n")
```

And here is the output:

```{r class.source=NULL, message=FALSE, warning=FALSE}
# Fitting the Model, Examining the Output

# Setting up input data
design_matrix <- as.data.frame(crashes_tracts_geo) %>%
  select(prop_pubtransit, median_income, median_age, frag_index, traffic) %>%
  mutate(median_income = log(median_income),
         traffic = log(traffic))
  
N <- nrow(crashes_tracts_geo)
y <- as.integer(crashes_tracts_geo$tot_crashes)
E <- as.integer(crashes_tracts_geo$population)
K <- ncol(design_matrix)
xs <- design_matrix
I <- length(unique(crashes_tracts_geo$tract_id))
i_index <- crashes_tracts_geo$tract_id
J <- length(unique(crashes_tracts_geo$time_id))
j_index <- crashes_tracts_geo$time_id

data_time_1 <- list(
  N = N,
  y = y,
  E = E,
  K = K,
  xs = xs,
  I = I,
  J = J,
  i_index = i_index,
  j_index = j_index
)

time_model_1 <- cmdstan_model(time_model_1_file)
set.seed(50)

time_fit_1 <- time_model_1$sample(data=data_time_1, 
                                  parallel_chains=cores, 
                                  refresh=0)

saveRDS(time_fit_1, file = "model-files/time_fit_1.rds")

print(time_fit_1)
```

My rhat values are all very close to 1.01 (the ideal being 1.00, the gold standard being anything below 1.01, and the "it's fine" standard being anything below 1.1). To be honest, I'm not super fussed with the actual coefficient values 

```{r class.source=NULL, message=FALSE, warning=FALSE}
# Posterior Checks
y_rep_time <- as_draws_matrix(time_fit_1$draws("y_rep"))
ppc_central_interval(y_rep_time, data_time_1$y)
```

```{r class.source=NULL, message=FALSE, warning=FALSE}
ppc_y_yrep_overlay(y_rep_time, data_time_1$y,
                     'Poisson model PPC\ny (blue dot) vs. y_rep (yellow 50% central interval, grey full extent)')
```

```{r class.source=NULL, message=FALSE, warning=FALSE}
# Check for Overdispersion
mean_y <- mean(data_time_1$y)
var_y <- var(data_time_1$y)

cat("Mean:", mean_y, "Variance:", var_y, "Ratio:", var_y / mean_y, "\n")
```

## Time Series Negative Binomial Model

```{r class.source=NULL, message=FALSE, warning=FALSE}
# The Stan Model
time_model_2_file = file.path('stan', 'poisson_time_2.stan')
cat(readLines(time_model_2_file), sep="\n")
```

```{r class.source=NULL, message=FALSE, warning=FALSE}
# Fitting the Model, Examining the Output

# Setting up input data
design_matrix <- as.data.frame(crashes_tracts_geo) %>%
  select(prop_pubtransit, median_income, median_age, frag_index, traffic) %>%
  mutate(median_income = log(median_income),
         traffic = log(traffic))
  
N <- nrow(crashes_tracts_geo)
y <- as.integer(crashes_tracts_geo$tot_crashes)
E <- as.integer(crashes_tracts_geo$population)
K <- ncol(design_matrix)
xs <- design_matrix
I <- length(unique(crashes_tracts_geo$tract_id))
i_index <- crashes_tracts_geo$tract_id
J <- length(unique(crashes_tracts_geo$time_id))
j_index <- crashes_tracts_geo$time_id

data_time_2 <- list(
  N = N,
  y = y,
  E = E,
  K = K,
  xs = xs,
  I = I,
  J = J,
  i_index = i_index,
  j_index = j_index
)

time_model_2 <- cmdstan_model(time_model_2_file)
set.seed(50)

time_fit_2 <- time_model_2$sample(data=data_time_2, 
                                  parallel_chains=cores, 
                                  refresh=0)

saveRDS(time_fit_2, file = "model-files/time_fit_2.rds")

time_summary <- time_fit_2$summary()
vars <- c('beta_intercept', 'beta0', 'betas[1]', 'betas[2]', 'betas[3]', 'betas[4]', 'betas[5]', 'sigma', 'rho', 'nb_disp')
time_summary_subset <- time_summary[time_summary$variable %in% vars, ]
numeric_cols <- sapply(time_summary_subset, is.numeric)
time_summary_subset[numeric_cols] <- round(time_summary_subset[numeric_cols], 3)

print(time_summary_subset)
```

```{r class.source=NULL, message=FALSE, warning=FALSE}
# Posterior Checks
y_rep_time2 <- as_draws_matrix(time_fit_2$draws("y_rep"))
ppc_central_interval(y_rep_time2, data_time_2$y)
```

```{r class.source=NULL, message=FALSE, warning=FALSE}
ppc_y_yrep_overlay(y_rep_time2, data_time_2$y,
                     'Poisson model PPC\ny (blue dot) vs. y_rep (yellow 50% central interval, grey full extent)')
```

```{r class.source=NULL, message=FALSE, warning=FALSE}
# Check Zero Inflation
n_zeros_obs <- sum(data_time_2$y == 0)
n_zeros_pred <- apply(y_rep_time2, 1, function(x) sum(x == 0))
mean_pred_zeros <- mean(n_zeros_pred)

print(paste("Actual N Zeros:", n_zeros_obs, "Mean Pred Zeros:", mean_pred_zeros))
```

# Combining BYM2 and Time Series Models

## Retrieving the Spatial Dependency Matrix

```{r class.source=NULL, message=FALSE, warning=FALSE}
# Get the spatial structure
tracts_uniques <- crashes_tracts_geo %>%
  select(tract_id, geometry, BoroCT2020) %>%
  distinct() %>%
  st_set_crs(st_crs(crashes_tracts_geo))

nyc_nbs = poly2nb(tracts_uniques)
nyc_coords = st_coordinates(st_centroid(tracts_uniques['geometry']))

# Create a spatial points sf object for centroids
centroids_sf <- st_as_sf(data.frame(x = nyc_coords[,1], y = nyc_coords[,2]), 
                         coords = c("x", "y"), 
                         crs = st_crs(tracts_uniques))

# Function to add a symmetric edge
add_symmetric_edge_nb <- function(nb, i, j) {
  nb_res = nb
  if (!(j %in% nb[[i]])) {
    if (nb[[i]][1] == 0) {
      nb_res[[i]] = j
    } else {
      nb_res[[i]] <- sort(c(nb[[i]], j))
    }
  }
  if (!(i %in% nb[[j]])) {
    if (nb[[j]][1] == 0) {
      nb_res[[j]] = i
     } else {
       nb_res[[j]] <- sort(c(nb[[j]], i))
     }
  }
  return(nb_res)
}

nyc_nbs_connected = add_symmetric_edge_nb(nyc_nbs, 
                                          which(tracts_uniques$BoroCT2020 == 3016400),
                                          which(tracts_uniques$BoroCT2020 == 5001800))
nyc_nbs_connected = add_symmetric_edge_nb(nyc_nbs_connected, 
                                          which(tracts_uniques$BoroCT2020 == 4088400),
                                          which(tracts_uniques$BoroCT2020 == 4107201))
nyc_nbs_connected = add_symmetric_edge_nb(nyc_nbs_connected, 
                                          which(tracts_uniques$BoroCT2020 == 4107201),
                                          which(tracts_uniques$BoroCT2020 == 4094201))
nyc_nbs_connected = add_symmetric_edge_nb(nyc_nbs_connected, 
                                          which(tracts_uniques$BoroCT2020 == 4099801),
                                          which(tracts_uniques$BoroCT2020 == 4103202))
nyc_nbs_connected = add_symmetric_edge_nb(nyc_nbs_connected, 
                                          which(tracts_uniques$BoroCT2020 == 2045600),
                                          which(tracts_uniques$BoroCT2020 == 2042600))

nyc_nbs_connected = add_symmetric_edge_nb(nyc_nbs_connected, 
                                          which(tracts_uniques$BoroCT2020 == 1029900),
                                          which(tracts_uniques$BoroCT2020 == 2026900))

lines_sp = nb2lines(nyc_nbs_connected, coords = nyc_coords, proj4string = st_crs(tracts_uniques)$wkt)
lines_sf = st_as_sf(lines_sp)

# Checking Work
spdep_map <- tm_shape(tracts_uniques) +
  tm_polygons(col = "white") + 
  tm_lines(col = "white", lwd = 2) + 

  # Add points for centroids
  tm_shape(centroids_sf) +
  tm_dots(size = 0.25, col = "black") +

  # Add lines for neighbors
  tm_shape(lines_sf) +
  tm_lines(col = "deepskyblue3")

spdep_map
```

```{r class.source=NULL, message=FALSE, warning=FALSE}
cat('is symmetric? ', is.symmetric.nb(nyc_nbs_connected), '\n')
summary(nyc_nbs_connected)
cat('nodes per component')
table(n.comp.nb(nyc_nbs_connected)$comp.id)
```

## Modeling

```{r class.source=NULL, message=FALSE, warning=FALSE}
# The Stan Model
full_model_1_file = file.path('stan', 'full_model_1.stan')
cat(readLines(full_model_1_file), sep="\n")
```

```{r class.source=NULL, message=FALSE, warning=FALSE}
# Setting up input data
nbs_adj <- nbs_to_adjlist(nyc_nbs_connected)

design_matrix <- as.data.frame(crashes_tracts_geo) %>%
  select(prop_pubtransit, median_income, median_age, frag_index, traffic) %>%
  mutate(median_income = log(median_income),
         traffic = log(traffic))
  
N <- nrow(crashes_tracts_geo)
y <- as.integer(crashes_tracts_geo$tot_crashes)
E <- as.integer(crashes_tracts_geo$population)
K <- ncol(design_matrix)
xs <- design_matrix
I <- length(unique(crashes_tracts_geo$tract_id))
i_index <- crashes_tracts_geo$tract_id
J <- length(unique(crashes_tracts_geo$time_id))
j_index <- crashes_tracts_geo$time_id
N_edges <- ncol(nbs_adj)
neighbors <- nbs_adj
tau <- get_scaling_factor(nyc_nbs_connected)

data_full_1 <- list(
  N = N,
  y = y,
  E = E,
  K = K,
  xs = xs,
  I = I,
  J = J,
  i_index = i_index,
  j_index = j_index,
  N_edges = N_edges,
  neighbors = neighbors,
  tau = tau
)

full_model_1 <- cmdstan_model(full_model_1_file)
set.seed(50)

full_fit_1 <- full_model_1$sample(data=data_full_1, 
                                  parallel_chains=cores, 
                                  refresh=0)

saveRDS(full_fit_1, file = "model-files/full_fit_1.rds")

full_summary <- full_fit_1$summary()
vars <- c('beta_intercept', 'beta0', 'betas[1]', 'betas[2]', 'betas[3]', 'betas[4]', 'betas[5]', 'sigma', 'rho', 'nb_disp')
full_summary_subset <- full_summary[full_summary$variable %in% vars, ]
numeric_cols <- sapply(full_summary_subset, is.numeric)
full_summary_subset[numeric_cols] <- round(full_summary_subset[numeric_cols], 3)

print(full_summary_subset)
```

```{r class.source=NULL, message=FALSE, warning=FALSE}
# Posterior Checks
y_rep_full <- as_draws_matrix(full_fit_1$draws("y_rep"))
ppc_central_interval(y_rep_full, data_full_1$y)
```

```{r class.source=NULL, message=FALSE, warning=FALSE}
ppc_y_yrep_overlay(y_rep_full, data_full_1$y,
                     'Poisson model PPC\ny (blue dot) vs. y_rep (yellow 50% central interval, grey full extent)')
```


# Intensive Diagnostic Steps

## Visualizing a Sample of Draws
```{r class.source=NULL, message=FALSE, warning=FALSE}
# Visualizing a sample of draws
mask <- sample(4000, 8)
index <- c("Draw 1", "Draw 2", "Draw 3", "Draw 4", "Draw 5", "Draw 6", "Draw 7", "Draw 8", "Original Sample")

y_rep_full_sample <- as.data.frame(y_rep_full[mask,]) %>%
  rbind(y)

y_rep_full_sample <- cbind(index, y_rep_full_sample) %>%
  pivot_longer(cols = -index)

density_check <- ggplot(data = y_rep_full_sample, aes(x = value)) + 
  geom_histogram(aes(y = after_stat(density)), 
                 bins = 100, 
                 color = "lightblue",
                 fill = "lightblue",
                 alpha = 0.75) + 
  theme_bw() + 
  labs(title = "Distribution Total Crashes per Census Tract (2024-25)",
       x = "Number of Crashes",
       y = "Density")

density_check + facet_wrap(~index, ncol = 3)
```

## Residuals by Time Period
```{r class.source=NULL, message=FALSE, warning=FALSE}
# Dealing with y_rep
y_actuals_full <- matrix(, nrow = 4000, ncol = length(y))

for (i in 1:4000){
  y_actuals_full[i, ] <- y
}

y_rep_full_matrix <- as.matrix(y_rep_full)
y_residuals_full <- y_actuals_full - y_rep_full_matrix

y_residuals_mean <- apply(y_residuals_full, 2, mean)
y_residuals_sd <- apply(y_residuals_full, 2, sd)

crashes_withresiduals <- cbind(crashes_tracts_geo, y_residuals_mean, y_residuals_sd)

# Creating the dataset
crashes_withresiduals_grouped <- crashes_withresiduals %>%
  group_by(monthyear) %>%
  summarize(mean_meanres = mean(y_residuals_mean), 
            mean_sdres = mean(y_residuals_sd),
            .groups = "drop") %>%
  st_drop_geometry() %>%
  mutate(
    monthyear_date = as.Date(paste0("01-", monthyear), format = "%d-%m-%y")
  )

# Time series plot of mean residuals
residuals_ts_boro_plot <- ggplot(data = crashes_withresiduals_grouped,
                       aes(x = monthyear_date,
                           y = mean_meanres)) + 
  geom_line() + 
  theme_minimal()

# Time series plot of sd residuals
residuals_ts_boro_plot_sd <- ggplot(data = crashes_withresiduals_grouped,
                       aes(x = monthyear_date,
                           y = mean_sdres)) + 
  geom_line() + 
  theme_minimal()

residuals_ts_boro_plot + residuals_ts_boro_plot_sd
```

## Residuals by Geography

```{r class.source=NULL, message=FALSE, warning=FALSE}
# Creating the dataset
crashes_withresiduals_grouped_geo <- crashes_withresiduals %>%
  group_by(BoroCT2020) %>%
  summarize(mean_meanres = mean(y_residuals_mean), 
            mean_sdres = mean(y_residuals_sd),
            .groups = "drop")

# Mapping mean and sd residuals
mean_residuals_map <- tm_shape(crashes_withresiduals_grouped_geo) + 
  tm_polygons(fill = "mean_meanres",
              fill.scale = tm_scale_intervals(values = "brewer.bu_gn",
                                              style = "jenks",
                                              midpoint = 0),
              fill.legend = tm_legend(title = "Average Mean Residuals by Tract"),
              lwd = 0.3) +
  tm_layout(legend.outside = TRUE, 
            legend.outside.position = "right")

sd_residuals_map <- tm_shape(crashes_withresiduals_grouped_geo) + 
  tm_polygons(fill = "mean_sdres",
              fill.scale = tm_scale_intervals(values = "brewer.blues",
                                              style = "jenks"),
              fill.legend = tm_legend(title = "Average SD of Residuals by Tract"),
              lwd = 0.3) +
  tm_layout(legend.outside = TRUE, 
            legend.outside.position = "right")

tmap_arrange(mean_residuals_map,
             sd_residuals_map,
             nrow = 1, ncol = 2)
```

NOTE THAT IF MORE TIME AND COMPUTING POWER, I WOULD LOOK TO MODEL VARYING DISPERSION BY BOROUGH

# Causal Inference via Difference-in-Difference

## Modeling

```{r class.source=NULL, message=FALSE, warning=FALSE}
# The Stan Model
did_model_1_file = file.path('stan', 'did_model_1.stan')
cat(readLines(did_model_1_file), sep="\n")
```

```{r class.source=NULL, message=FALSE, warning=FALSE}
# Setting up input data
nbs_adj <- nbs_to_adjlist(nyc_nbs_connected)

design_matrix <- as.data.frame(crashes_tracts_geo) %>%
  select(prop_pubtransit, median_income, median_age, frag_index, traffic,
         cp_zone, after_cp) %>%
  mutate(median_income = log(median_income),
         traffic = log(traffic))
  
N <- nrow(crashes_tracts_geo)
y <- as.integer(crashes_tracts_geo$tot_crashes)
E <- as.integer(crashes_tracts_geo$population)
K <- ncol(design_matrix)
xs <- design_matrix
I <- length(unique(crashes_tracts_geo$tract_id))
i_index <- crashes_tracts_geo$tract_id
J <- length(unique(crashes_tracts_geo$time_id))
j_index <- crashes_tracts_geo$time_id
N_edges <- ncol(nbs_adj)
neighbors <- nbs_adj
tau <- get_scaling_factor(nyc_nbs_connected)
did <- crashes_tracts_geo$treatment

data_did_1 <- list(
  N = N,
  y = y,
  E = E,
  K = K,
  xs = xs,
  I = I,
  J = J,
  i_index = i_index,
  j_index = j_index,
  N_edges = N_edges,
  neighbors = neighbors,
  tau = tau,
  did = did
)

did_model_1 <- cmdstan_model(did_model_1_file)
set.seed(50)

did_fit_1 <- did_model_1$sample(data=data_did_1, 
                                  parallel_chains=cores, 
                                  refresh=0)

saveRDS(did_fit_1, file = "model-files/did_fit_1.rds")

did_summary <- did_fit_1$summary()
vars <- c('theta', 'beta_intercept', 'beta0', 'betas[1]', 'betas[2]', 'betas[3]', 'betas[4]', 'betas[5]', 'sigma', 'rho', 'nb_disp')
did_summary_subset <- did_summary[did_summary$variable %in% vars, ]
numeric_cols <- sapply(did_summary_subset, is.numeric)
did_summary_subset[numeric_cols] <- round(did_summary_subset[numeric_cols], 3)

print(did_summary_subset)
```




